Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-21 paper/paper.tex
Revise section 2

\scw{Need to explain the notations in the text. What are these operations?
What do readers need to understand about them? Help me understand the figure}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & \ottkw{v}_{[[n]] + [[i]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

\scw{motivate parallel reduction. Why are you telling us this now. Is this
an explanation of definitional equality and its properties?}
Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
from? I'm not familiar with it. Maybe you can replace it with ``definitional
equivalence'' as that is how it is used in \lang?}
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests\scw{but you prove it below?} it is an equivalence relation.

We sketch out some key properties about parallel reduction and
coherence without giving their proofs. Our technique for establishing
those results is based on \citet{takahashi-parallel-reduction} and a
modern exposition of the same technique can be found in
\citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to derive that coherence is
exactly the same as untyped $\beta$-equivalence.\scw{cite barendregt?}

@@

\begin{figure}[ht]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, Eq, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
\scw{Is the precondition for T-ABS really admissible? Isn't it required
to ensure predicativity? Have you proved it? Have you proved regularity?}
\yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
  was formulated correctly (which wasn't the case in the version
  of the paper you read). I revised \rref{T-Univ} so the premise is
  not admissible }
\scw{Do you have a formation rule for identity types?}
\yl{Added}

In \rref{T-Var}, $[[G i]]$ is the partial function defined
with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
i ]] := [[G i]] $.  \scw{explain the renaming in the conclusion of
this lemma}
The precondition $[[i < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
i]]$ has a defined output.

\Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. It is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types). % and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.
\scw{Maybe move more of this paragraph and the next to the discussion/related
  work section, and add a forward pointer here.}
\scw{Just say that like Coq, the definition of the system uses untyped
equality}

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity.
Section~\ref{sec:extension} explains how we
generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

Finally, since our system has an infinite universe hierarchy, we can
present the system à la Russell by using the same judgment form $[[G
|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
is no need to distinguish between big types and small types
and duplicate our typing specification.
\scw{Is there an advantage for Tarski universes even with infinite
hierarchy? or no?}

@@

\section{Logical Relation}
\scw{This section pre-supposes that we want to define a logical relation,
but doesn't precisely state what we want to use it to prove, and why a logical
relation is suitable. (And why the property you want to prove is difficult to
show!) Should add more motivation here.}

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}).
Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
The logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Because the type of $I$ depends on $i$, I'd prefer a syntax that
keeps them in order. }
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

@@

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\scw{I didn't understand this figure without reading the text below.}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

@@

\begin{corollary}[Logical Consistency]
  \label{corollary:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{corollary}
\begin{proof}
  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
\end{proof}

>>>>

\scw{Need to explain the notations in the text. What are these operations?
What do readers need to understand about them? Help me understand the figure}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

\scw{motivate parallel reduction. Why are you telling us this now. Is this
  an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par}. Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, Eq, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}
% Figure~\ref{fig:par} shows the definition of the parallel reduction
% relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
% represent its transitive and reflexive closure, which in turn allows us to define
% the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
% from? I'm not familiar with it. Maybe you can replace it with ``definitional
% equivalence'' as that is how it is used in \lang?}
% We say that two terms $[[a]]$
% and $[[b]]$ are coherent if they can eventually reduce to some common
% term $[[c]]$ through parallel reduction. The symmetric notation of
% coherence suggests\scw{but you prove it below?} it is an equivalence relation.

% We sketch out some key properties about parallel reduction and
% coherence without giving their proofs. Our technique for establishing
% those results is based on

% \scw{cite barendregt?}
% \yl{resolved}

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
% \scw{Is the precondition for T-ABS really admissible? Isn't it required
% to ensure predicativity? Have you proved it? Have you proved regularity?}
% \yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
%   was formulated correctly (which wasn't the case in the version
%   of the paper you read). I revised \rref{T-Univ} so the premise is
%   not admissible }
% \scw{Do you have a formation rule for identity types?}
% \yl{Added}

In \rref{T-Var}, $[[G n]]$ is the partial function defined
by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
n ]] := [[G n]]$.
The precondition $[[n < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
scoped differently. In the typing judgment $[[G |- a : A]]$, we want
both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
type is to the head of $[[G]]$, the smaller de Bruijn it has since
there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
the full context $[[G]]$ rather than a truncated version of $[[G]]$.

\Rref{T-Conv} uses the convertibility relation from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion is reminiscent of Barendregt's Pure Type
Systems~\citet{barendregt1991introduction} and makes our formulation different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
takes the form $\Gamma \vdash a \equiv b : A$, from which one can
usually derive  $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving
subject reduction.

\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. This assures us that we do not lose much generality working
with a system with untyped conversion. We include a detailed
discussion in Section~\ref{sec:relatedwork} on how type-directed and
untyped equality affect the applicability of our technique.

% As a result, there is no loss of generality from

% It is trivial to embed a system with judgmental
% equality to a system with untyped equality by erasing typing
% information. As a result, it is easy to port our consistency result from our
% type system to a variant with typed judgmental equality
% as long as the typed system does not include $\eta$ laws
% that would require type annotations (e.g. the $\eta$-law for unit
% types).
% and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.
\yl{Unresolved: is it ok to say our system's conversion rule is untyped like
Coq?}
% \scw{Maybe move more of this paragraph and the next to the discussion/related
%   work section, and add a forward pointer here.}
% \scw{Just say that like Coq, the definition of the system uses untyped
% equality}

% Working with a system with untyped equality has the huge benefit that
% the confluence result for untyped parallel reduction
% (Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
% resort to the complex syntatic (resp. semantic) technique from
% \citet{siles2012pure} (resp. \citet{decagda}) to resolve
% the circularity of subject reduction and $\Pi$-injectivity.
% Section~\ref{sec:extension} explains how we
% generalize our technique to include $\eta$-law for functions and
% show the existence of normal form for well-typed (open and closed)
% terms, achieving a similar level of expressiveness of the type system
% and strength of metatheoretic property as \citet{decagda}.

% Finally, since our system has an infinite universe hierarchy, we can
% present the system à la Russell by using the same judgment form
%$[[G |- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
% is no need to distinguish between big types and small types
% and duplicate our typing specification.
% \scw{Is there an advantage for Tarski universes even with infinite
% hierarchy? or no?}

@@

\section{Logical Relation}
% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
Before we define our logical relation, we first formally specify the
consistency property that we want to prove.
\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is similarly defined as a type that has no term. A related
property, referred to as the termination property (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.

A naive attempt to proving Theorem~\ref{theorem:consistency} by
induction on the derivation $[[empty |- a : Void]]$ would succeed at
almost all cases except for \rref{T-App}. In the application
case, we are given $[[empty |- b : Pi A B]]$ and $[[empty |- a : A]]$, and
the equality that $[[B {a} = Bool]]$. Our goal is to show that
$[[empty |- b a : Void]]$ is not possible. However, note that there is
nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
because neither $[[Pi A B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
We have no way of deriving a contradiction from $[[empty |- b a :
Void]]$. The takeaway from this failed attempt is that, in order to
derive the consistency, we need to know something about types other
than $[[Void]]$. From a pragmatic point of view, proof by logical
relation can be seen as a sophisticated way of strengthening the
induction hypothesis. From the strengthened property, the fundamental
theorem, we will be able to derive consistency as a corollary.

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Themetavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

@@

To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\scw{I didn't understand this figure without reading the text below.}
\end{figure}

@@

\begin{proof}
  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
\end{proof}
